---
title: "Motif Analysis"
author: "Amanda Charbonneau"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: "cerulean"
    number_sections: true
    toc: true
    toc_depth: 5
    toc_float: true
    collapsed: false
    df_print: paged
    code_folding: hide
params:
  fimo: "anxiety.meta.full.fs.tbl.bed/fimo.tsv"
  cons: "anxiety.meta.full.fs.tbl.bed/anxiety.meta.full.fs.tbl.bed.cons"
  bed: "anxiety.meta.full.fs.tbl.bed/anxiety.meta.full.fs.tbl.bed"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = "/Users/amanda/BrainTF")
# Install function for packages    
packages<-function(x){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x,repos="http://cran.r-project.org")
    require(x,character.only=TRUE)
  }
}

packages( plyr )
packages( dplyr ) #munging
packages(knitr)
packages(DT)

#packages( plotrix ) #shading
#packages( cowplot ) #ggplot gridding
#packages( lme4 ) #modeling
#packages( ggthemes )
#packages( ggplot2 ) #plotting
#packages( reshape2 ) #munging
#packages( car ) #anova


```

```{r globalvariables}

qval1 <- .1
qval2 <- .05

```

# Setup `r params$fimo`

I wrote some code to extract the full set of human transcription factor motifs from [CIS-BP]( http://cisbp.ccbr.utoronto.ca) and [turn them into a format useable by MEME](https://github.com/ACharbonneau/TFpsych/blob/master/1_format_motifs.sh). I then [took the results from a number of GWAS studies on human psychiatric diseases, extracted the SNPs with a p value of less than 1e-7, and searched a 200bp window centered on each SNP for all the TF motifs](https://github.com/ACharbonneau/TFpsych/blob/master/2_makebedfiles_scanmotifs.sh)

This script takes the output of that pipeline and extracts hit counts.


```{r GetData, warning=FALSE}

sessionInfo()


fimofile <- read.table(params$fimo, sep = "\t", header = T)
fimofile$sequence_name <- as.character(fimofile$sequence_name)

bedfile <- read.table(params$bed)
colnames(bedfile) <- c("chr", "snpM100", "snpP100", "snpid", "pval")
bedfile$sequence_name <- paste(bedfile$chr, ":", bedfile$snpM100, "-", bedfile$snpP100, sep="")

unannotatedsnps <- read.table("RawData/ConsPeaks_FilterByAll_LDSR.txt", col.names = "snpid")
unannotatedsnps$conservation <- "UnAnnConserved"

TFnames <- read.table("RawData/TF_Information.txt", sep = "\t", fill = TRUE, header = TRUE)
TFnames$motif <- substr(as.character(TFnames$Motif_ID),start = 1, nchar(as.character(TFnames$Motif_ID))-5)

```

```{r JoinData, warning=FALSE}
idjoin <- left_join(fimofile, bedfile)

alljoin <- left_join(idjoin, unannotatedsnps)

alljoin$snpbase <- alljoin$snpM100 + 100

# is snp in a motif?

alljoin$snpinmotif <- ifelse(alljoin$start <= 100 & alljoin$stop >= 100, "Yes", "No")

# is snp Conserved?
alljoin$conservation <- ifelse(is.na(alljoin$conservation), "UnAnnNonConserved", "UnAnnConserved")


```


## Filtering

```{r filter}

filtereddata <- alljoin %>% filter(q.value < qval1 ) %>% group_by(snpid, sequence_name, start, stop, motif_id, conservation, snpinmotif) %>% summarise(n=n())


```
First, I've filtered all data to have a qvalue of less than `r qval1` and removed perfectly overlapping (palindrome) motif hits. 

A [second filtering step](https://github.com/ACharbonneau/ubiquitous-succotash/blob/master/2.0_metafiltering.sb) removes SNPs that were identified as being part of an LDSR catagory. SNPs are only retained if they were in the list of SNPs identified as conserved in the 30 primates data or in Conserved_LindbladToh and were also NOT in any of:

 * Coding_UCSC
 * Promoter_UCSC
 * UTR_3_UCSC
 * UTR_5_UCSC
 * lincRNA
 * miRNA
 * rRNA
 * TssAFlnk

The above SNP list has `r length(unannotatedsnps$snpid)` entries.

This filtering results in `r length(alljoin$snpid)` motif matches of `r length(unique(alljoin$sequence_name))` SNPs.


```{r testmotifhits}
#motifs with n=2 were palindrome hits

filteredbysnp <- alljoin %>% filter(q.value < qval1 ) %>% group_by(snpid, conservation) %>% summarise(n=n())

filteredbysnp$At_Least_5_Motifs <- ifelse(filteredbysnp$n >= 5, "Yes", "No")

filteredbysnp$At_Least_10_Motifs <- ifelse(filteredbysnp$n >= 10, "Yes", "No")

filteredbysnp$At_Least_20_Motifs <- ifelse(filteredbysnp$n >= 20, "Yes", "No")

filteredbysnp$At_Least_50_Motifs <- ifelse(filteredbysnp$n >= 50, "Yes", "No")



```

## Testing motif hit numbers, qvalue of `r qval1`

Our hypothesis is that conserved SNPs will have more surrounding motifs than non-conserved SNPs, and that they are conserved because they are as of yet unidentified enhancers. 

I've taken the above data table and counted the number of motifs that match to a given SNP window and tallied them. Below are contingency tables for various count cutoffs. 

A column set labeled 'At_least_5_Motifs' 'Yes' is the number of SNPs in the dataset that had five or more motifs hit in their 200bp window. 

### 5 motif cutoff 
```{r}


ftable(filteredbysnp[,c(2,4)])

```
### 10 motif cutoff 

```{r}
ftable(filteredbysnp[,c(2,5)])

```

### 20 motif cutoff 

```{r}
ftable(filteredbysnp[,c(2,6)])
```

### 50 motif cutoff 

```{r}

ftable(filteredbysnp[,c(2,7)])
```

## Testing motif hit numbers, qvalue of `r qval2`

This is the same type of data as the last section, however the qvalue filtering has been shifted to `r qval2`

```{r stringenttestmotifhits}
#motifs with n=2 were palindrome hits

stringentfilteredbysnp <- alljoin %>% filter(q.value < qval2 ) %>% group_by(snpid, conservation) %>% summarise(n=n())


stringentfilteredbysnp$At_Least_5_Motifs <- ifelse(stringentfilteredbysnp$n >= 5, "Yes", "No")

stringentfilteredbysnp$At_Least_10_Motifs <- ifelse(stringentfilteredbysnp$n >= 10, "Yes", "No")

stringentfilteredbysnp$At_Least_20_Motifs <- ifelse(stringentfilteredbysnp$n >= 20, "Yes", "No")

stringentfilteredbysnp$At_Least_50_Motifs <- ifelse(stringentfilteredbysnp$n >= 50, "Yes", "No")



```

### 5 motif cutoff 
```{r}


ftable(stringentfilteredbysnp[,c(2,4)])

```
### 10 motif cutoff 

```{r}
ftable(stringentfilteredbysnp[,c(2,5)])

```

### 20 motif cutoff 

```{r}
ftable(stringentfilteredbysnp[,c(2,6)])
```

### 50 motif cutoff 

```{r}

ftable(stringentfilteredbysnp[,c(2,7)])
```


## TF frequencies

```{r}
motifcounts <- group_by(filtereddata, motif_id ) %>% summarise(n=n())

motifsnamed <- left_join(motifcounts, TFnames, by=c("motif_id" = "motif")) %>% select(motif_id, DBID, TF_Name, Family_Name, Motif_Type, n)

datatable(motifsnamed)
```


